Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-14 _tex/manuscript.tex
Built site for gh-pages

\caption{}\label{T_ded44}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

\caption{}\label{T_f573d}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

@@

\caption{}\label{T_eeac7}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

@@

In table~\ref{tbl-data-unknowns} we show the combinations of the midline
extension status and state of the contralateral LNL II. Using only
available information, we would get to 6 out of 29 patients for the
prevalence of contralateral LNL II invovlement among the early
T-category patients with midline extension. But the column with the
``unknown'' numbers suggests that the true prevalence could range from 6
out of 29 + 59 (7\%) up to 12 out of 29 + 6 (34\%). Assuming that every
one of the possible combinations has the same a priori probability
(uniform prior) and accouting for permutations via the binomial factor,
the posterior marginalized over the unknowns could be written as

\[
\begin{aligned}
p(x; \alpha=1, \beta=1) &= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell} (1-x)^{n+m - (k+\ell)} \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta \\
&= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell+1} (1-x)^{n+m - (k+\ell) + 1}
\end{aligned}
\]

@@

The distribution that marginalizes over the unknown midline extension
cases is shifted towards a lower prevalence. This makes intuitive sense:
With only six patients among the 65 with unknown midline extension
status, the upper bound for the prevalence has moved down compared to
six out of 29.

>>>>

\caption{}\label{T_42bd2}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

\caption{}\label{T_c1c15}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

@@

\caption{}\label{T_78884}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

@@

In table~\ref{tbl-data-unknowns} we show the combinations of the midline
extension status and state of the contralateral LNL II. Using only
available information, we would get to 6 out of 29 patients for the
prevalence of contralateral LNL II invovlement among the early
T-category patients with midline extension. But the column with the
``unknown'' numbers suggests that the true prevalence could range from 6
out of 29 + 59 (7\%) up to 12 out of 29 + 6 (34\%). We assume that every
one of the possible combinations has the same a priori probability
(uniform prior) but we do need to account for permutations via the
binomial factor. In general, let us say we have \(n\) patients with
midline extension, of which \(k\) harbor metastases in the contralateral
LNL II. And we have \(m\) patients with unknown midline extension, of
which \(\ell\) presented with diasease in contralateral LNL II, then the
posterior distribution - marginalized over the unknown data - over the
true prevalence \(x \in \left[ 0, 1 \right]\) is given by:

\[
p(x \mid n, k, m, \ell, \alpha, \beta) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \frac{P(i, j \mid x )}{\binom{n}{k}} \right] \cdot P(n, k \mid x) \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta
\]

Note the two binomial likelihoods: one being marginalized over and one
acting implicitly as a prior to the summed terms. This role of the
likelihood over the data with known midline extension as a prior to the
part where that data is missing corrects for the fact that there are
highly unlikely combinations among those we marginalize over. E.g.,
expecting that exactly those 6 patients with contralateral LNL II
involvemnt are also coincidentially the only ones of the 65 with a tumor
crossing the mid-sagittal line is very unlikely. This is accounted for
by the fact that this possibility has only one permutation and it is
weakened by the likelihood over the data with known midline extension.
With a uniform prior beta distribution (\(\alpha=1\) and \(\beta=1\)),
the above simplifies to:

\[
p(x \mid n, k, m, \ell) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \binom{n+i}{k+j} x^j (1 - x)^{i-j} \right] \cdot x^{k+1} (1 - x)^{n-k+1}
\]

@@

The distribution that marginalizes over the unknown midline extension
cases is shifted towards a lower prevalence. This makes intuitive sense:
With only 6 patients among the 65 with unknown midline extension status,
the upper bound for the prevalence has moved down compared to 6 out of
29.
